---
title: "Week 3 - Homework"
author: "STAT 420, Summer 2019, BALAJI SATHYAMURTHY (BALAJIS2)"
date: ''
output:
  word_document:
    toc: yes
  pdf_document: default
  html_document:
    toc: yes
urlcolor: cyan
---

# Directions

- Be sure to remove this section if you use this `.Rmd` file as a template.
- You may leave the questions in your final document.

***

## Exercise 1 (Using `lm` for Inference)

For this exercise we will use the `cats` dataset from the `MASS` package. You should use `?cats` to learn about the background of this dataset.

**(a)** Fit the following simple linear regression model in `R`. Use heart weight as the response and body weight as the predictor. 

\[
Y_i = \beta_0 + \beta_1 x_i + \epsilon_i
\]

```{r}
library(MASS)
cat_model = lm(Hwt~Bwt,data=cats)
```

Store the results in a variable called `cat_model`. Use a $t$ test to test the significance of the regression. Report the following:

```{r}
tstatistic = summary(cat_model)$coefficients["Bwt","t value"]
pvalue = summary(cat_model)$coefficients["Bwt","Pr(>|t|)"]
alpha = 0.05
pvalue < alpha
```

- The null and alternate hypotheses \[H_0: \beta_1=0 \ H_1: \beta_1\neq0\]
- The value of the test statistic is `r tstatistic`

- The p-value of the test is `r pvalue`

- A statistical decision at $\alpha = 0.05$

    `r pvalue` is less than $\alpha = 0.05$, we reject the null hypothesis \[H_0:  \beta_1=0\]

- A conclusion in the context of the problem
  
  Since the null hypothesis is rejected, there must a linear relationship between   the heart weight and body weight in the CATS data set  

When reporting these, you should explicitly state them in your document, not assume that a reader will find and interpret them from a large block of `R` output.

**(b)** Calculate a 90% confidence interval for $\beta_1$. Give an interpretation of the interval in the context of the problem.

```{r}
lower_confint_beta1 = confint(cat_model,level=0.90)[2,1]
upper_confint_beta1 = confint(cat_model,level=0.90)[2,2]
```

We are 90% confident with an increase in body weight by 1 kg the heart weight will be between `r lower_confint_beta1` and `r upper_confint_beta1`

**(c)** Calculate a 99% confidence interval for $\beta_0$. Give an interpretation of the interval in the context of the problem.

```{r}
lower_confint_beta0 = confint(cat_model,level=0.99)[1,1]
upper_confint_beta0 = confint(cat_model,level=0.99)[1,2]
```

We are 99% confident that for the body weight of 0 kg the heart weight will be between `r lower_confint_beta0` and `r upper_confint_beta0`

**(d)** Use a 99% confidence interval to estimate the mean heart weight for body weights of 2.1 and 2.8 kilograms. Which of the two intervals is wider? Why?

```{r}
new_data = data.frame(Bwt = c(2.1,2.8))

a = predict(cat_model,newdata = new_data,interval = c("confidence"),level = 0.99)[1,2]
b = predict(cat_model,newdata = new_data,interval = c("confidence"),level = 0.99)[1,3]
c = predict(cat_model,newdata = new_data,interval = c("confidence"),level = 0.99)[2,2]
d = predict(cat_model,newdata = new_data,interval = c("confidence"),level = 0.99)[2,3]
```

We are 99% confident that the mean heart weight of body weight of 2.1kg is between `r a` and `r b` and the heart weight for body weight of 2.8 kg is between `r c` and `r d`. The interval for 2.1kg `r b-a` is wider than for 2.8kg `r d-c` 

**(e)** Use a 99% prediction interval to predict the heart weight for body weights of 2.8 and 4.2 kilograms.

```{r}
new_data1 = data.frame(Bwt = c(2.8,4.2))

e = predict(cat_model,newdata = new_data1,interval = c("prediction"),level = 0.99)[1,2]
f = predict(cat_model,newdata = new_data1,interval = c("prediction"),level = 0.99)[1,3]
g = predict(cat_model,newdata = new_data1,interval = c("prediction"),level = 0.99)[2,2]
h = predict(cat_model,newdata = new_data1,interval = c("prediction"),level = 0.99)[2,3]
```

We are 99% confident that a new observation for heart weight for a body weight of 2.8kg will fall between `r e` and `r f` and for heart weight of 4.2kg will fall between `r g` and `r h`.

**(f)** Create a scatterplot of the data. Add the regression line, 90% confidence bands, and 90% prediction bands.

```{r echo=FALSE}
bwt_grid = seq(min(cats$Bwt),max(cats$Bwt),by=0.01)
bwt_ci_band = predict(cat_model,newdata=data.frame(Bwt=bwt_grid),interval = "confidence",level = 0.90)
bwt_pi_band = predict(cat_model,newdata=data.frame(Bwt=bwt_grid),interval = "prediction",level = 0.90)

plot(Hwt ~ Bwt,data=cats,xlab="body weight",ylab="Heart weight",main="Cats heart weight vs body weight",col = "blue")
abline(cat_model,col="orange",lwd=4)

lines(bwt_grid,bwt_ci_band[,2],col ="red",lwd = 3)
lines(bwt_grid,bwt_ci_band[,3],col ="red",lwd = 3)
lines(bwt_grid,bwt_pi_band[,2],col ="green",lwd = 3)
lines(bwt_grid,bwt_pi_band[,3],col ="green",lwd = 3)

```

**(g)** Use a $t$ test to test:

- $H_0: \beta_1 = 4$
- $H_1: \beta_1 \neq 4$

Report the following:

- The value of the test statistic
```{r}
beta1_est = summary(cat_model)$coefficients["Bwt","Estimate"]
std_error = summary(cat_model)$coefficients["Bwt","Std. Error"]
beta1_hyp = 4
test_statistic = (beta1_est-beta1_hyp)/std_error
test_statistic
```
The value of test statistic is `r test_statistic`

- The p-value of the test

```{r}
p_value = 2 * pt(abs(test_statistic),df=nrow(cats)-2,lower.tail = FALSE)
```

The p-value of the test is `r p_value`

- A statistical decision at $\alpha = 0.05$

```{r}
alpha_val = 0.05
p_value < alpha_val
```

We fail to reject the null hypothesis since p-value `r p_value` is less than $\alpha = 0.05$

When reporting these, you should explicitly state them in your document, not assume that a reader will find and interpret them from a large block of `R` output.

***

## Exercise 2 (More `lm` for Inference)

For this exercise we will use the `Ozone` dataset from the `mlbench` package. You should use `?Ozone` to learn about the background of this dataset. You may need to install the `mlbench` package. If you do so, do not include code to install the package in your `R` Markdown document.

For simplicity, we will re-perform the data cleaning done in the previous homework.

```{r}
data(Ozone, package = "mlbench")
Ozone = Ozone[, c(4, 6, 7, 8)]
colnames(Ozone) = c("ozone", "wind", "humidity", "temp")
Ozone = Ozone[complete.cases(Ozone), ]
```

**(a)** Fit the following simple linear regression model in `R`. Use the ozone measurement as the response and wind speed as the predictor. 

\[
Y_i = \beta_0 + \beta_1 x_i + \epsilon_i
\]

```{r}
ozone_wind_model = lm(ozone~wind,data=Ozone)
t_value_wind =summary(ozone_wind_model)$coefficients["wind","t value"]
p_value_wind =summary(ozone_wind_model)$coefficients["wind","Pr(>|t|)"]
alpha_val_wind = 0.01
p_value_wind < alpha_val_wind
```

Store the results in a variable called `ozone_wind_model`. Use a $t$ test to test the significance of the regression. Report the following:

- The null and alternative hypotheses
  
  \[H_0: \beta_1=0 \ H_1: \beta_1\neq0\]
  
- The value of the test statistic is `r t_value_wind`

- The p-value of the test is `r p_value_wind`

- A statistical decision at $\alpha = 0.01$

  Since p-value `r p_value_wind` is not less than $\alpha = 0.01$, we failed to reject   the null hypothesis
  
- A conclusion in the context of the problem
  
  Since we failed to reject the null hypothesis it's possible that there is no linear relationship between ozone measurement and wind speed

When reporting these, you should explicitly state them in your document, not assume that a reader will find and interpret them from a large block of `R` output.

**(b)** Fit the following simple linear regression model in `R`. Use the ozone measurement as the response and temperature as the predictor. 

\[
Y_i = \beta_0 + \beta_1 x_i + \epsilon_i
\]

Store the results in a variable called `ozone_temp_model`. Use a $t$ test to test the significance of the regression. Report the following:

```{r}
ozone_temp_model = lm(ozone~temp,data=Ozone)
t_value_temp =summary(ozone_temp_model)$coefficients["temp","t value"]
p_value_temp =summary(ozone_temp_model)$coefficients["temp","Pr(>|t|)"]
alpha_val_temp = 0.01
p_value_temp < alpha_val_temp
```

- The null and alternative hypotheses

  \[H_0: \beta_1=0 \ H_1: \beta_1\neq0\]
  
- The value of the test statistic is `r t_value_temp`

- The p-value of the test is `r p_value_temp`

- A statistical decision at $\alpha = 0.01$

  Since p-value `r p_value_temp` is less than $\alpha = 0.01$, we reject the null 
  hypothesis
  
- A conclusion in the context of the problem
  
  Since we rejected the null hypothesis there is a linear relationship between ozone measurement and temperature  

When reporting these, you should explicitly state them in your document, not assume that a reader will find and interpret them from a large block of `R` output.

***

## Exercise 3 (Simulating Sampling Distributions)

For this exercise we will simulate data from the following model:

\[
Y_i = \beta_0 + \beta_1 x_i + \epsilon_i
\]

Where $\epsilon_i \sim N(0, \sigma^2).$ Also, the parameters are known to be:

- $\beta_0 = -5$
- $\beta_1 = 3.25$
- $\sigma^2 = 16$

We will use samples of size $n = 50$.

**(a)** Simulate this model $2000$ times. Each time use `lm()` to fit a simple linear regression model, then store the value of $\hat{\beta}_0$ and $\hat{\beta}_1$. Set a seed using **your** birthday before performing the simulation. Note, we are simulating the $x$ values once, and then they remain fixed for the remainder of the exercise.

```{r}
birthday = 18760613
set.seed(birthday)
n = 50
x = seq(0, 10, length = n)
```

```{r}
beta_hat_0 = rep(0,2000)
beta_hat_1 = rep(0,2000)

for (i in seq_along(beta_hat_1)) {
  x = runif(n = 50, 0, 10)
  num_obs = 50
  beta_0 = -5
  beta_1 = 3.25
  sigma  = sqrt(16)
  epsilon = rnorm(n = num_obs,mean = 0, sd = sigma)
  y = beta_0 + beta_1 * x + epsilon
  sim_fit = lm(y ~ x)
  beta_hat_0[i] = coef(sim_fit)[1]  
  beta_hat_1[i] = coef(sim_fit)[2]
}

```

**(b)** Create a table that summarizes the results of the simulations. The table should have two columns, one for $\hat{\beta}_0$ and one for $\hat{\beta}_1$. The table should have four rows:

- A row for the true expected value given the known values of $x$
```{r}
beta_0
beta_1
```
  
- A row for the mean of the simulated values
```{r}
mean_sim_beta0=mean(beta_hat_0)
mean_sim_beta1=mean(beta_hat_1)
```
- A row for the true standard deviation given the known values of $x$

```{r}
Sxx = sum( (x-mean(x)) ^ 2 )
var_beta0_hat = sigma ^ 2 * (1/n + mean(x)^2/Sxx)
var_beta1_hat = sigma ^ 2 / Sxx
sd_beta0_hat = sqrt(var_beta0_hat)
sd_beta1_hat = sqrt(var_beta1_hat)
```
- A row for the standard deviation of the simulated values

```{r}
sd_sim_beta0=sd(beta_hat_0)
sd_sim_beta1=sd(beta_hat_1)
```

```{r}
Values = c("True expected value","Mean of simulated values","True SD of actual values","SD of simulated values")
Beta0values = c(beta_0,mean_sim_beta0,sd_beta0_hat,sd_sim_beta0)
Beta1values = c(beta_1,mean_sim_beta1,sd_beta1_hat,sd_sim_beta1)

output = data.frame(Value = Values,beta0 = Beta0values,beta1 = Beta1values)

library(flextable)
library(magrittr)
flextable(output)
```

**(c)** Plot two histograms side-by-side:

- A histogram of your simulated values for $\hat{\beta}_0$. Add the normal curve for the true sampling distribution of $\hat{\beta}_0$.

```{r}
hist(beta_hat_0,col="green",prob="TRUE",breaks=20,main="beta_hat_0 distribution")
curve(dnorm(x,beta_0,sd_beta0_hat),add=TRUE,lwd=3,col="red")
```

- A histogram of your simulated values for $\hat{\beta}_1$. Add the normal curve for the true sampling distribution of $\hat{\beta}_1$.

```{r}
hist(beta_hat_1,col="green",prob="TRUE",breaks=20,main="beta_hat_1 distribution")
curve(dnorm(x,beta_1,sd_beta1_hat),add=TRUE,lwd=3,col="red")
```

***

## Exercise 4 (Simulating Confidence Intervals)

For this exercise we will simulate data from the following model:

\[
Y_i = \beta_0 + \beta_1 x_i + \epsilon_i
\]

Where $\epsilon_i \sim N(0, \sigma^2).$ Also, the parameters are known to be:

- $\beta_0 = 5$
- $\beta_1 = 2$
- $\sigma^2 = 9$

We will use samples of size $n = 25$.

Our goal here is to use simulation to verify that the confidence intervals really do have their stated confidence level. Do **not** use the `confint()` function for this entire exercise.

**(a)** Simulate this model $2500$ times. Each time use `lm()` to fit a simple linear regression model, then store the value of $\hat{\beta}_1$ and $s_e$. Set a seed using **your** birthday before performing the simulation. Note, we are simulating the $x$ values once, and then they remain fixed for the remainder of the exercise.

```{r}
birthday = 18760613
set.seed(birthday)
n = 25
x = seq(0, 2.5, length = n)
```


```{r}
beta_hat_1 = rep(0,2500)
s_e        = rep(0,2500)

for (i in seq_along(beta_hat_1)) {
  x = runif(n = 25, 0, 2.5)
  num_obs = 25
  beta_0 = 5
  beta_1 = 2
  sigma  = sqrt(9)
  epsilon = rnorm(n = num_obs,mean = 0, sd = sigma)
  y = beta_0 + beta_1 * x + epsilon
  sim_fit = lm(y ~ x)
  beta_hat_1[i] = coef(sim_fit)[2]
  s_e[i]        = summary((lm(y~x)))$sigma
}

```

**(b)** For each of the $\hat{\beta}_1$ that you simulated, calculate a 95% confidence interval. Store the lower limits in a vector `lower_95` and the upper limits in a vector `upper_95`. Some hints:
- You will need to use `qt()` to calculate the critical value, which will be the same for each interval.
- Remember that `x` is fixed, so $S_{xx}$ will be the same for each interval.
- You could, but do not need to write a `for` loop. Remember vectorized operations.

```{r}
crit_value_95 =abs( qt((1-0.95)/2, df = n-2) ) 
Sxx           =sum((x- mean(x))^2)
lower_95 = beta_hat_1 - crit_value_95 * s_e / sqrt(Sxx)
upper_95 = beta_hat_1 + crit_value_95  *s_e / sqrt(Sxx)
```

**(c)** What proportion of these intervals contains the true value of $\beta_1$?

```{r}
cnt = 0
for (i in seq_along(lower_95)) {
if (lower_95[i] < beta_1 & beta_1 < upper_95[i]){ 
  cnt = cnt + 1}
}
total_proportion_95 = cnt/length(lower_95)
```

Around `r total_proportion_95` of the values contains true value of $\beta_1$

**(d)** Based on these intervals, what proportion of the simulations would reject the test $H_0: \beta_1 = 0$ vs $H_1: \beta_1 \neq 0$ at $\alpha = 0.05$?

```{r}
cnt = 0
for (i in seq_along(lower_95)) {
if (lower_95[i] < 0 & 0 < upper_95[i]){ 
  cnt = cnt + 1}
}
total_proportion_0_95 = 1- (cnt/length(lower_95))
```

Around `r total_proportion_0_95` of the values will be rejected for null hypothesis $H_0: \beta_1 = 0$

**(e)** For each of the $\hat{\beta}_1$ that you simulated, calculate a 99% confidence interval. Store the lower limits in a vector `lower_99` and the upper limits in a vector `upper_99`.

```{r}
crit_value_99 =abs( qt((1-0.99)/2, df = n-2) ) 
Sxx           =sum((x- mean(x))^2)
lower_99 = beta_hat_1 - crit_value_99 * s_e / sqrt(Sxx)
upper_99 = beta_hat_1 + crit_value_99  *s_e / sqrt(Sxx)
```

**(f)** What proportion of these intervals contains the true value of $\beta_1$?

```{r}
cnt = 0
for (i in seq_along(lower_99)) {
if (lower_99[i] < beta_1 & beta_1 < upper_99[i]){ 
  cnt = cnt + 1}
}
total_proportion_99 = cnt/length(lower_99)
```

Around `r total_proportion_99` of the values contains true value of $\beta_1$

**(g)** Based on these intervals, what proportion of the simulations would reject the test $H_0: \beta_1 = 0$ vs $H_1: \beta_1 \neq 0$ at $\alpha = 0.01$?

```{r}
cnt = 0
for (i in seq_along(lower_99)) {
if (lower_99[i] < 0 & 0 < upper_99[i]){ 
  cnt = cnt + 1}
}
total_proportion_0_99 = 1- (cnt/length(lower_99))
```

Around `r total_proportion_0_99` of the values will be rejected for null hypothesis $H_0: \beta_1 = 0$

***

## Exercise 5 (Prediction Intervals "without" `predict`)

Write a function named `calc_pred_int` that performs calculates prediction intervals:

$$
\hat{y}(x) \pm t_{\alpha/2, n - 2} \cdot s_e\sqrt{1 + \frac{1}{n}+\frac{(x-\bar{x})^2}{S_{xx}}}.
$$

for the linear model

$$
Y_i = \beta_0 + \beta_1 x_i + \epsilon_i.
$$

**(a)** Write this function. You may use the `predict()` function, but you may **not** supply a value for the `level` argument of `predict()`. (You can certainly use `predict()` any way you would like in order to check your work.)

The function should take three inputs:

- `model`, a model object that is the result of fitting the SLR model with `lm()`
- `newdata`, a data frame with a single observation (row)
    - This data frame will need to have a variable (column) with the same name as the data used to fit `model`.
- `level`, the level (0.90, 0.95, etc) for the interval with a default value of `0.95`

The function should return a named vector with three elements:

- `estimate`, the midpoint of the interval
- `lower`, the lower bound of the interval
- `upper`, the upper bound of the interval

```{r}
calc_pred_int = function(model,newdata,level=0.95){
  
  est_value_95 = predict(model,newdata,interval = "prediction")[1,1]
  lower_value_95 = predict(model,newdata,interval = "prediction")[1,2]
  upper_value_95 = predict(model,newdata,interval = "prediction")[1,3]
  
  est_value_for_parm = (level * est_value_95) / 0.95
  lower_value_for_parm = (level * lower_value_95) / 0.95
  upper_value_for_parm = (level * upper_value_95) / 0.95
  
  c(estimate = est_value_for_parm, 
    lower = lower_value_for_parm, 
    upper = upper_value_for_parm)
}

```

**(b)** After writing the function, run this code:

```{r}
newcat_1 = data.frame(Bwt = 4.0)
calc_pred_int(cat_model, newcat_1)
```

**(c)** After writing the function, run this code:

```{r}
newcat_2 = data.frame(Bwt = 3.3)
calc_pred_int(cat_model, newcat_2, level = 0.99)
```


